Higher-Order unification for free!

### 2024-05-03 paper/main.tex
minor

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
When it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),a},
    {X,\lambda x.a}}}
  {{{A,a},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected,
in turn breaking \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
% decompilation phase.
To address this issue, we post-process the compiler's output
to ensure the following property:

\begin{proposition}[\mapStore is a bijection]
  After compilation, for each \Fo{}-variable $X$ in \foUnifPb{} and for each
  \Ho{}-variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
  (for an arity $n$).
\end{proposition}

\noindent
Note that the execution in \Ho{} may require allocating new variables as
explained in~\cref{sec:decompilation}. The property above does not
concern these variables since they are fresh and have no corresponding
variable in \Fo.

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\wellb_{\beta \eta}(\foUnifPb)$,
  if $\ \foUnifPb$ contains the same \Fo{} variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\def\chain{\ensuremath{\mathcal{C}}\xspace}

\begin{proof}[Proof sketch]
  Let $X$ a \Fo variable appearing in two different subterms $s_1$ and $s_2$. in
  $s_1$, $X$ is applied to the list of terms $x_1\dots x_m$ and, in $s_2$, it is
  applied to $y_1\dots y_n$. Here, $m < n$. By \mapdeduplication, the
  compilation genereates two distinct \Ho variables for $X$, say $A$ and $C$,
  related by a chain \chain of \linketa{}s, with length $n - m$. We prove that
  if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
  these terms fails to unify. By \progressetaleft, the assignment of $A$
  activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
  chain collapses and unification is performed between $\lambda x_n\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
  we work with \wellb terms, this unification succeeds iff it does in \Fo. If,
  otherwise, $p < n$, then two cases should be considered: $t_1$ is a rigid
  term, then, $t_2$ must be an eta expansion of $t_1$, implying $q = 0$. If it
  is not the case, then the successive progressions made by \progressetaleft
  would eventually cause a failure. If, on the other hand, $t_1$ is a
  unification variable, \elpiIn{scope-check} guarantees that $t_2$ does not use
  $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
  After this check, $m - n - p$ links of \chain remain suspended, as we expect:
  decompilation is charged wake all links in \linkStore and eventually relate
  $t_1$ with $t_2$.
  %
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
  % 
  % Nella nuova versione sottintendo questo caso: p = 0 e t_1 è flessibile
  %
  % Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  % $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  % term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  % so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  % Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  % sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  % links is shrinked by $q$. Note that if this length is bigger then $n-m$
  % the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  % Otherwise,
  % if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  % with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  % due to the constraint of the number of lambda by which $C^n$ should start.
  %
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
  %
  % If $A^m$ is instantiated, 
  %
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}

@@

\noindent
Note that $\mapping{X}{C}{0}$ desappears from \mapStore
in favour of the auxiliary \linketa \linketaM{}{C}{\lambda x.B_{x}}.
The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
$\linkStore_2$ by \progressetaleft, assigning <<$a \appsep x$>> to $B_x$.
The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
turn triggers $\linkStore_1$ by \progressetaleft. This link perfroms
$\lambda x.a \Ue \lambda x.(a \appsep x)$, which, as expected, fails.

@@

\noindent
In order to support this scenario we have to improve a little our compiler
and \progress routines.
In particular $\hoUnifPb_1$ generates $\sigma = \{ A
\mapsto \lambda x. a\}$ making $\sigma\hoUnifPb_2$ equal to
$(\lambda x. a) \appsep a \Ue a$ that \Ue{} cannot solve since it
lacks rules \ref{clause:beta1} and \ref{clause:beta2}.

To address this problem the compiler must recognize and replace
\notllambda terms with fresh variables and generate a new kind of links
that we call \linkbeta.

@@

\begin{definition}[\progBetaFail]
  \progress fails whenever:
  \begin{enumerate}
  \item
  there exists another link $l' \in \linkStore$ with the same \lhs as $l$; or
  \item the \lhs of $l$
  become rigid.
  \end{enumerate}
  \label{def:progBetaFail}
\end{definition}
\noindent
We relax this condition and accommodate for
% the implementation of 
heuristics in~\ref{sec:heuristics}.

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 
\begin{itemize}
  \item $\testmaybeeta~T$ does not hold (anymore). In this case $X$ is unified
        with $T$
  \item $T$ $\eta$-contracts to a term $T'$ not starting with the \elpiIn{lam}
        constructor. In this case $X$ is unified with $T'$
\end{itemize}
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $\foUnifPb_i$ be the first problem in \foUnifPb, such that it is not in \llambda and le
  $\sigma$ be the substitution obtained solving
  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of \hstep
  for the $i-1$ time,
  each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 
  unification problem, is instantiated if it is known for sure
  that $X$ is no more a \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
  % each variable that the
  % compiler used to replace a ``problematic'' term is assigned and hence
  by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
  is solved and removed. Note that, in this case, all
  calls to \Ue{} are between terms in \llambda and by \cref{th:fidelity-beta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,
  then by \cref{def:progBetaFail} unification fails as the
  corresponding unification in \Fo{} would.
\end{proof}

>>>>

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
When it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),a},
    {X,\lambda x.a}}}
  {{{A,a},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected,
in turn breaking \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
% decompilation phase.
To address this issue, we post-process \mapStore
to ensure the following property:

\begin{proposition}[\mapStore is a bijection]
  After compilation, for each \Fo{}-variable $X$ in \foUnifPb{} and for each
  \Ho{}-variable $A$ in \hoUnifPb{} there is exactly one entry $X \mapsto A^n$
  in \mapStore.
\end{proposition}

\noindent
Note that running \Ue{} may require allocating new variables in \Ho{} as
explained in~\cref{sec:decompilation}. The property above ignores these variables since they are fresh and have no corresponding
variable in \Fo.

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\wellb_{\beta \eta}(\foUnifPb)$,
  if $\ \foUnifPb$ contains the same \Fo{}-variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\def\chain{\ensuremath{\vec{c}}\xspace}

\begin{proof}[Proof sketch]
  Let $X$ be a \Fo{}-variable appearing in two different subterms $s_1$ and $s_2$.
  In $s_1$ variable $X$ is applied to the list of terms $x_1\dots x_m$ while, in $s_2$,
  it is applied to $y_1\dots y_n$. Without loss of generality we take $m < n$.
  After \mapdeduplication we have two distinct \Ho{}-variables for $X$, say $A$ and $C$,
  related by a chain \chain of \linketa{}s of length $n - m$. We prove that
  if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
  these terms fails to unify.\todo{a unification fails if they fail to unify is a tautology, to be clarified} By \progressetaleft, the assignment of $A$
  activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
  chain collapses and unification is performed between $\lambda x_n\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
  we work with \wellb terms, this unification succeeds iff it does in \Fo.
  Otherwise if $p < n$ then we have to consider two cases. 1) If $t_1$ is a rigid
  term then $t_2$ must be the $\eta$-expansion of $t_1$, implying $q = 0$. If it
  is not the case, then \progressetaleft
  would eventually cause a failure. 2) If $t_1$ is a
  unification variable then \elpiIn{scope-check} guarantees that $t_2$ does not use
  $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
  After this check, $m - n - p$ links of \chain remain suspended, as we expect:
  decompilation is charged to wake all links in \linkStore and eventually relate
  $t_1$ with $t_2$.
  %
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
  % 
  % Nella nuova versione sottintendo questo caso: p = 0 e t_1 è flessibile
  %
  % Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  % $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  % term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  % so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  % Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  % sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  % links is shrinked by $q$. Note that if this length is bigger then $n-m$
  % the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  % Otherwise,
  % if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  % with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  % due to the constraint of the number of lambda by which $C^n$ should start.
  %
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
  %
  % If $A^m$ is instantiated, 
  %
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}

@@

\noindent
Note that $\mapping{X}{C}{0}$ desappears from \mapStore
in favour of the auxiliary \linketa \linketaM{}{C}{\lambda x.B_{x}}.
The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
$\linkStore_2$ by \progressetaleft, assigning <<$a \appsep x$>> to $B_x$.
The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
turn triggers $\linkStore_1$ by \progressetaleft. In turn, this performs
$\lambda x.a \Ue \lambda x.(a \appsep x)$ that fails as expected.

@@

\noindent
In order to support this scenario we have to improve a little our compiler
and \progress routines.
In particular $\hoUnifPb_1$ generates $\sigma = \{~ A
\mapsto \lambda x. a ~\}$ making $\sigma\hoUnifPb_2$ equal to
$(\lambda x. a) \appsep a \Ue a$ that \Ue{} cannot solve since it
lacks rules \ref{clause:beta1} and \ref{clause:beta2}.

To address this problem the compiler must recognize
\notllambda terms
and replace them
with fresh variables and generate a new kind of links
that we call \linkbeta.

@@

\begin{definition}[\progBetaFail]
  \progress fails when either
  \begin{itemize}
  \item
  there exists another link $l' \in \linkStore$ with the same \lhs as $l$; or
  \item the \lhs of $l$
  become rigid.
  \end{itemize}
  \label{def:progBetaFail}
\end{definition}
\noindent
We relax this condition and accommodate for
% the implementation of 
heuristics in~\ref{sec:heuristics}.

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 
\begin{itemize}
  \item $\testmaybeeta~T$ does not hold (anymore), in this case $X$ is unified
        with $T$; or
  \item $T$ $\eta$-contracts to a term $T'$ not starting with the \elpiIn{lam}
        constructor. In this case $X$ is unified with $T'$
\end{itemize}
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $\foUnifPb_i$ be the first problem in \foUnifPb{} such that it is not in \llambda and let
  $\sigma$ be the substitution obtained solving
  $\hoUnifPb_1\ldots\hoUnifPb_{i-1}$. After the execution of \hstep
  for the $i-1$ time,
  each variable $X$ corresponding to a \maybeeta subterm in the original \Fo 
  unification problem, is instantiated iff it is known for sure
  that $X$ is no more in \maybeeta. If $\sigma\hoUnifPb_i$ is in \llambda, then 
  % each variable that the
  % compiler used to replace a ``problematic'' term is assigned and hence
  by \cref{def:progBetaLL,def:progBetaRH} the associated \linkbeta
  is solved and removed. Note that, in this case, all
  calls to \Ue{} are between terms in \llambda and by \cref{th:fidelity-beta}
  fidelity is guaranteed. If $\sigma\hoUnifPb$ is still in \notllambda,
  then by \cref{def:progBetaFail} unification fails as the
  corresponding unification in \Fo{} would.
\end{proof}
